---
title: "Task 4 - Purple Urchins at Naples Reef in 2010"
author: "Taylor Medina"
date: "10/23/2021"
output: html_document
---

This task requires the wrangling of the `urchins.csv` data set to include on purple urchins in the Naples Reef Sampling Site in the year 2010. Then, a statistical summary is needed in order to visualize the difference in urchin size in two different treatments, one with no kelp removal and the other with annual kelp removal. A histogram and a qq-plot is generated to visualize the differences in purple urchin sizes.

**Data:** Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 

### Setup
```{r setup, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)

# load in packages
library(tidyverse)
library(here)
library(kableExtra)
library(plotly)
library(janitor)
library(lubridate)

```

### Data Wrangling
The `lubridate::ymd()` is used to convert the date column to a date class in order to splice out the year to be filtered out. The commands `group_by()` and `summarize()` is used to generate a statistical summary of the sizes of the purple urchins in different treatments.
```{r, warning = FALSE, message = FALSE}

napl_urchins <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names() %>% 
  mutate(date = ymd(date), # change format and class of the date column
         year = year(date)) %>% # add a new column that splices the year from the date column
  filter(year == "2010", 
         common_name == "Purple Urchin",
         site == "NAPL")

napl_urchins_stats <- napl_urchins %>% 
  group_by(treatment) %>% 
  summarize(mean_diameter_cm = mean(size), # calculated mean diameter
            standard_deviation = sd(size), # calculates standard deviation of size
            sample_size = n())

```

### Histogram Plot
```{r}

napl_urchins_hist <- ggplot(data = napl_urchins,
                            aes(x = size)) +
  geom_histogram(aes(fill = treatment), color = "black", binwidth = 0.5) + # make color based on treatment
  facet_wrap(~ treatment) + # split up plot by treatment
  xlim(0,7) + # change the x-limit
  scale_y_continuous(expand = c(0,0)) + # make x-axis start at 0

  theme_classic() + 
  labs(x = "diameter (cm)",
       y = "count")
  
ggplotly(napl_urchins_hist)

```
**Figure 1:** Size Distribution of Purple Urchins in different Treatments

### QQ-Plot
```{r}

napl_urchins_qq <- ggplot(data = napl_urchins,
                          aes(sample = size)) +
  stat_qq() + # plots points
  stat_qq_line(aes(color = treatment)) + # adds trend line to qq-plot
  facet_wrap(~ treatment) + # splits up plot by treatment
  theme_classic() +
  labs(x = "standard deviation (cm)",
       y = "Size Quartile")

ggplotly(napl_urchins_qq)

```
**Figure 2:** QQ-plot of the Purple Urchin sizes in different Treatments

The purple urchins in the control treatment follow a normal distribution as compared to that in the annual treatment where the sizes are skewed right, according to **Figure 1**, where they tend to be smaller. **Figure 2** displays how the quartiles fro the control treatment are fitted better to the line better than the annual treatment, where the majority of the sizes are on the lower quartiles.

### Statistical Table
```{r}

napl_urchins_stats %>% 
  kable(col.names = c("Treatment",
                      "Mean Diameter (cm)",
                      "Standard Deviation (cm)",
                      "Sample Size"),
        caption = "Table 1: Statistical Summary of Purple Urchin Diameters (cm) ") %>% 
  kable_classic(full_width = FALSE) %>% 
  footnote(general = "Annual: annual kelp removal. Control: no kelp removal") # adds a footnote to the table

```

