---
title: "Task 3 - Santa Barbara Urchins"
author: "Taylor Medina"
date: "10/21/2021"
output: html_document
---

This task requires the use of sea urchin data from 5 Santa Barbara reefs to generate statistical summaries of the size distribution, population change over time, and common statistical measures such as mean, median, and standard deviation.

**Data:** Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 

### **Setup**
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 2) # sets sig figs to 2

# Load in packages
library(tidyverse)
library(here)
library(kableExtra)
library(plotly)
library(janitor)
library(lubridate)

```

### **Section 1**
The goal for this section is to generate a histogram comparing the distribution of urchin diameters among red and purple urchins and to generate a summary table of the mean urchin diameter, the median urchin diameter, the standard deviation, and the sample size of the purple and red urchin populations. The `geom_histogram()` and `kable()` functions are used.

#### Creating a histogram plot
```{r, message = FALSE}

urchins <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names()

urchins_plot <- ggplot(data = urchins,
       aes(x = size)) +
  geom_histogram(bins = 10, aes(fill = common_name), color = "black") +
  facet_wrap(~ common_name) + # splits plots by species
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("purple","red")) + # specifies colors manually
  theme_classic() +
  labs(x = "urchin diameter (cm)",
       y = "frequency",
       fill = "Common Name",
       title = "Size Distribution of Sea Urchins in Santa Barbara Reefs") # sets name of legend

ggplotly(urchins_plot)

```
**Figure 1:** Size Distribution of Purple and Red Sea Urchins in Santa Barbara Reefs

#### Generating a summary table
```{r}

urchin_stats <- urchins %>% 
  group_by(common_name) %>% 
  summarize(mean_diameter_cm = mean(size, na.rm = TRUE), # calculates mean diameter
            median_diameter_cm = median(size, na.rm = TRUE), # calculates median diameter
            standard_deviation = sd(size, na.rm = TRUE), # calculates standard deviation 
            sample_size = n()) # sample size count

urchin_stats %>% 
  kable(col.names = c("Species", 
                      "Mean Diameter (cm)",
                      "Median Diameter (cm)",
                      "Standard Deviation (cm)",
                      "Sample Size"),
        caption = "Table 1: Summary Statistics of Purple and Red Urchins in Santa Barbara Reefs") %>% 
  kable_classic(full_width = FALSE)

```

The red urchins are 2.31 cm larger on average (or 52% larger) than purple urchins in Santa Barbara reefs according to **Table 1**. They also have a median diameter that is 2.5 cm greater and a larger standard deviation as well. According to **Figure 1**, the purple urchins are more localized along the smaller diameters while the red urchins are more evenly spread out.

### **Section 2**
The goal for this section is to manipulate the 'urchins.csv' data set to only include purple urchins and change the site names to their full names. Then, a line plot needs to be generated to compare urchin populations by sample site for each year they were sampled. The class of the date in the dataset is changed using `mdy()` and the `case_when()` function is used for the *for* statements to change the site names to their full name. 

#### Data Wrangling
```{r, message = FALSE}

purple_urchins <- urchins %>% 
  filter(common_name == "Purple Urchin") %>% 
  mutate(date = mdy(date), # converts the date column class from function to date
         year = year(date), # makes a new column splicing the year from the date column
         site_full = case_when( # makes a new column converting the short names to the full names of the sites
           site == "NAPL" ~ "Naples Reef",
           site == "AQUE" ~ "Arroyo Quemado Reef",
           site == "CARP" ~ "Carpinteria Reef",
           site == "MOHK" ~ "Mohawk Reef",
           site == "IVEE" ~ "Isla Vista Reef"))

purple_urchins_stats <- purple_urchins %>% 
  group_by(year,site_full) %>% 
  summarize(count = n()) # generates a count of purple observations by year and site

```

#### Generating a line plot
```{r}

purple_urchins_plot <- ggplot(data = purple_urchins_stats,
                              aes(x = year, y = count)) +
  geom_point(aes(color = site_full)) +
  geom_line(aes(color = site_full)) +
  theme_classic() +
  guides(col = guide_legend("Sampling Sites")) + # changes name of legend
  labs(x = "year",
       y = "sea urchin count",,
       title = "Sea urchin Populations in 5 Santa Barbara Reefs") 

ggplotly(purple_urchins_plot)

```
**Figure 2:** Comparison of Purple Urchin Counts in 5 Santa Barbara Reefs

Overall, there is a decline in the population of sea urchins in the Santa Barbara Reef. To determine if there is actual decline, the sample site areas need to be kept constant and the yearly sampling needs to be done at the same time as jumps in population could be associated with breeding season or natural moments of population decline due to the season. 